---
title: "Predicting Protein levels using leveles of mRNA in Tissue Samples"
author: "Abdullah Farouk"
date: '2018-04-08'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load Libraries
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readxl)
```

# Read Data in 
```{r}
mRNA <- read_excel('tpm.xlsx')

protein_copy_number <- read_excel('protein_copy_numbers.xlsx')

conversion_factor <- read_excel('conversion_factor.xlsx')

```


# Remove the first 4 columns, corresponding to the order they are displayed in Figure 2A
```{r}
fig_cols <- c('HEK293/Kidney',	'HepG2/Liver', 'Lung/A549',	'Breast/MCF7')

protein_copy_number <- protein_copy_number %>% 
  select(-one_of(fig_cols))

dim(protein_copy_number) == dim(mRNA)

conversion_factor <- conversion_factor %>% 
  select(-one_of(c('X__1', 'X__2')))


```



# Convert protein counts and conversion factor datasets into a numeric arguement
```{r, warning=FALSE, message=FALSE}
protein_copy_number[,2:21] <- sapply(protein_copy_number[,2:21], function(x) as.numeric(as.character(x)))

conversion_factor[,2:22] <- sapply(conversion_factor[,2:22], function(x) as.numeric(as.character(x)))

#Convert genes into factors
mRNA$Gene <- mRNA$Gene %>% as.factor()

protein_copy_number$Gene <- protein_copy_number$Gene %>% as.factor()

colnames(conversion_factor)[1] <- 'Gene'

conversion_factor$Gene <- conversion_factor$Gene %>% as.factor()

str(mRNA)

str(protein_copy_number)

str(conversion_factor)
```


# Convert data into long form (for graphing and mixed effects model fitting)
```{r}
conversion_factor_long <- gather(conversion_factor[,-2], key = 'Tissue', value = 'RTP', -'Gene')

protein_copy_number_long <- gather(protein_copy_number, key = 'Tissue', value = 'Protein_copy_number', -'Gene')

#Convert protein copy numbers to actual levels
protein_copy_number_long$Protein_copy_number <- protein_copy_number_long$Protein_copy_number*1000

mRNA_long <- gather(mRNA, key = 'Tissue', value = 'TPM', -'Gene')

#Arrange data by Gene
conversion_factor_long <- conversion_factor_long %>% arrange(Gene)

protein_copy_number_long <- protein_copy_number_long %>% arrange(Gene)

mRNA_long <- mRNA_long %>% arrange(Gene)

```



# EDA
```{r, warning=FALSE, message=FALSE}
#Check if conversion factor is constant across tissues
set.seed(10)

gene_levels <- conversion_factor$Gene[sample(length(conversion_factor$Gene), 3)] %>% as.character()

one <- conversion_factor_long %>% 
  filter(Gene %in% gene_levels) 

one <- one %>% 
  ggplot(aes(x = Tissue, y = RTP, colour = Tissue)) + 
  geom_point() +
  labs(title = "mRNA To Protein (RTP) Ratio for 3 Ranomly Selected Genes", x = "Tissue/Cell Line", y = "RTP") + 
  facet_wrap(~Gene, scales = 'free') +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), plot.title = element_text(hjust = 0.5), legend.position = 'none')

ggsave(one, filename = 'onegene.png')

bp <- ggplot(conversion_factor_long, aes(x = Tissue, y = RTP, colour = Tissue)) + 
  geom_boxplot() + 
  labs(title = "mRNA To Protein (RTP) Ratio across Tissue/Cell Lines", x = "Tissue/Cell Line", y = "RTP") + 
  theme_light() +
  scale_fill_brewer(palette="Dark2") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1), plot.title = element_text(hjust = 0.5), legend.position = 'none')

bp
ggsave(bp, file="boxplot.png")
```


# Construct dataset to fit mixed effects model
```{r}
data_comb <- conversion_factor_long

#Calculate median RTP values by leaving that specific tissue out
gene_levels <- data_comb$Gene


for(i in 1:length(gene_levels)) {
  
    x <- data_comb %>% filter(Gene == gene_levels[i])
    
    data_comb$medianrtp[i] <- median(x$RTP[-i], na.rm = TRUE)
    
}

#Divide median RTP by 1000 to get sensible predictions
data_comb$medianrtp <- data_comb$medianrtp

data_comb$TPM <- mRNA_long$TPM

#Convert protein copy numbers to actual levels
data_comb$protein <- protein_copy_number_long$Protein_copy_number

data_comb$predicted_protein <- data_comb$medianrtp*data_comb$TPM

```



# Look at whether a random slope and intercept is needed for each gene
```{r, warning=FALSE, message=FALSE}
set.seed(9)

# Plot protein vs MRNA values for 8 randomly selected Genes
gene_names <- conversion_factor$Gene[sample(length(conversion_factor$Gene), 9)] %>% as.character()

plot_data <- data_comb %>% filter(Gene %in% gene_names)

fp <- plot_data %>% 
  ggplot(aes(x = TPM, y = protein, group = Gene, colour = Gene)) + 
  geom_point() + 
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE) + 
  facet_wrap(~Gene, scales = 'free') + 
  labs(title = "Protein vs mRNA levels for 9 Randomly selected Genes", x = "mRNA Levels", y = "Protein Levels") + 
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), plot.title = element_text(hjust = 0.5), legend.position = 'none')

fp
ggsave(fp, file="facetplot.png")
```


# Fit a mixed effects model
```{r}
library('lme4')

data_comb_scaled <- data_comb

data_comb_scaled[,5:7] <- scale(data_comb[, 5:7], center = TRUE, scale = TRUE)

fit <- lmer(protein ~ TPM + (1+TPM|Gene), data = data_comb_scaled, REML=FALSE) #Add REML=FALSE when performing likelihood ratio tests
summary(fit)
```


# Interpret results from the model
```{r}
#Reduced model
null_fit <- lmer(protein ~ (1+TPM|Gene), data = data_comb_scaled, REML=FALSE)

anova(fit, null_fit)
```

# Check correlations between predicted and observed protein levels
```{r, message=FALSE, warning=FALSE}
#Check if conversion factor is constant across tissues
set.seed(13)

gene_levels <- data_comb$Gene[sample(length(data_comb$Gene), 6)] %>% as.character()

comp <- data_comb %>% 
  filter(Gene %in% gene_levels) 

comp <- comp %>% 
  ggplot(aes(x = predicted_protein, y = protein, colour = Gene)) + 
  geom_point() +
  labs(title = "Correlation between Observed and Predicted Protein levels", x = "Predicted Protein Levels", x = "Predicted Protein Levels", y = "Observed Protein Levels") + 
  facet_wrap(~Gene, scales = 'free') +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), plot.title = element_text(hjust = 0.5), legend.position = 'none') +
  annotate("segment", x=-Inf, xend=Inf,y=-Inf, yend=Inf, linetype = 2)

ggsave(comp, filename = 'comp.png')

# pp <- data_comb %>% 
#   ggplot(aes(x = predicted_protein, y = protein, colour = Gene, group = Gene)) +
#   geom_point() +
#   geom_jitter() +
#   labs(title = "Correlation between Observed and Predicted Protein levels", x = "Predicted Protein Levels", y = "Observed Protein Levels") +
#   theme_light()+
#   theme(legend.position = 'none', plot.title = element_text(hjust = 0.5))
# 
# pp
# ggsave(pp, file="predictioncomp.png")

#Check correlations
gene_levels <- levels(data_comb$Gene)

correlations <- c()

for(i in 1:length(gene_levels)) {
  
    x <- data_comb %>% filter(Gene %in% gene_levels[i])
    
    correlations[i] <- cor(x = x$predicted_protein, y = x$protein, use = 'everything', method = 'spearman')
    
}

w <- na.omit(correlations)


cor(x = data_comb$protein, y = data_comb$TPM, use = 'complete.obs', method = 'spearman')
```


# Check if protein levels vary across genes and across Tissues
```{r}
bp_tissue <- ggplot(data_comb, aes(x = Tissue, y = protein, colour = Tissue, group = Tissue)) + 
  geom_boxplot() + 
  labs(title = "Protein levels across Tissue/Cell Lines", x = "Tissue/Cell Line", y = "Protein Levels") + 
  theme_light() +
  scale_fill_brewer(palette="Dark2") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1), plot.title = element_text(hjust = 0.5), legend.position = 'none')

bp_tissue
ggsave(bp_tissue, file="boxplottissue.png")

#Now check for genes
bp_gene <- ggplot(data_comb, aes(x = Gene, y = protein, colour = Gene)) + 
  geom_boxplot() + 
  labs(title = "Protein levels across Genes", x = "Gene", y = "Protein Levels") + 
  theme_light() +
  scale_fill_brewer(palette="Dark2") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1), plot.title = element_text(hjust = 0.5), legend.position = 'none')

bp_gene
#ggsave(bp_gene, file="boxplotgene.png")
```


# Check correlations between mRNA and protein levels across genes
```{r}
# plot(conversion_factor[5,3:22])
#Check if conversion factor is constant across tissue samples
# boxplot(conversion_factor[,3:22])
# 
# hist(as.numeric(t(conversion_factor[ ,2])))
# 
# conversion_factor_complete <- na.omit(conversion_factor)
# 
# hist(as.numeric(conversion_factor_complete[sample(1:17,1), 3:22]))
# 
# #Check if mrna and protein levels are normally distributed
# hist(as.numeric(mRNA[3, ]))
# 
# hist(as.numeric(protein_copy_number[sample(1:55,1), ]))
# 
# #check to see how gene-specific conversion factors are computed
# rtp_factor <- protein_copy_number$A431 * 1000 / mRNA$A431
# 
# protein_complete <- na.omit(protein_copy_number)
# 
# #Find these genes in the mRNA dataframe
# 
# mrna_complete <- mRNA[which(mRNA$Gene %in% protein_complete$Gene), ]
# 
# #Try and verify findings in the article
# spearman_corr <- cor(x = mrna_complete[, 2:21], y = protein_complete[, 2:21], method = 'spearman')
# 
# pearson_corr <- cor(x = mrna_complete[, 2:21], y = protein_complete[, 2:21], method = "pearson")
# 
# #Compute mean correlations across genes
# avg_spearman_corr <- colMeans(spearman_corr)

```

# Visualize correlations
```{r}
# protein_and_rna <- cbind(mRNA, protein_copy_number[, 2:21])
# 
# protein_and_rna <- na.omit(protein_and_rna)
# 
# plot(x = protein_and_rna[, 2], y = protein_and_rna[, 22])
# 
# protein_and_rna %>% 
#   ggplot(aes(x = protein_and_rna[, 2], y = protein_and_rna[, 22], colour = protein_and_rna[,1])) +
#   geom_point() +
#   xlab("mRNA levels") + ylab("Protein Copy levels")
# 
# cor(x = protein_and_rna[, 2], y = protein_and_rna[, 22], method = "spearman")

```


# Visualize correlations across tissues
```{r}
#Have Genes along the columns
# protein_complete_1 <- t(protein_copy_number)
# 
# protein_complete_1 <- protein_complete_1[-1, ]
# 
# protein_complete_1 <- mapply(protein_complete_1, FUN=as.numeric)
# 
# protein_complete_1 <- matrix(data = protein_complete_1, ncol = 55, nrow = 20)
# 
# #Label its columns and rows
# colnames(protein_complete_1) <- protein_copy_number$Gene
# 
# rownames(protein_complete_1) <- colnames(protein_copy_number[,2:21])
# 
# str(protein_complete_1)
# 
# #Do the same with mrna data
# mrna_complete_1 <- t(mRNA)
# 
# mrna_complete_1 <- mrna_complete_1[-1, ]
# 
# mrna_complete_1 <- mapply(mrna_complete_1, FUN=as.numeric)
# 
# mrna_complete_1 <- matrix(data = mrna_complete_1, ncol = 55, nrow = 20)

#Label its columns and rows
# colnames(mrna_complete_1) <- mRNA$Gene
# 
# rownames(mrna_complete_1) <- colnames(mRNA[,2:21])
# 
# str(mrna_complete_1)
# 
# #Do the same with conversion factors
# rtp_1 <- conversion_factor[,-2]
# 
# rtp_1 <- t(conversion_factor)
# 
# rtp_1 <- rtp_1[-1, ]
# 
# rtp_1 <- mapply(rtp_1, FUN=as.numeric)
# 
# rtp_1 <- matrix(data = rtp_1, ncol = 55, nrow = 20)

#Label its columns and rows
# colnames(rtp_1) <- conversion_factor$Gene
# 
# rownames(rtp_1) <- colnames(conversion_factor[,3:22])
# 
# str(rtp_1)
# 
# boxplot(rtp_1)
```



# Check for correlations across tissues now
```{r}
#Try and verify findings in the article
# spearman_corr_gene <- cor(x = mrna_complete_1, y = protein_complete_1, method = 'spearman')
# 
# pearson_corr_gene <- cor(x = mrna_complete_1, y = protein_complete_1, method = "pearson")
# 
# average_spearman_corr_across_tissues <- colMeans(spearman_corr_gene)
# 
# average_pearson_corr_across_tissues <- colMeans(pearson_corr_gene)
# 
# #Visualize them
# protein_and_rna_1 <- as.data.frame(cbind(mrna_complete_1, protein_complete_1))
# 
# protein_and_rna_1 %>% 
#   ggplot(aes(x = protein_and_rna_1[, 1], y = protein_and_rna_1[, 18], colour = rownames(protein_and_rna_1))) +
#   geom_point() +
#   xlab("Gene ANXA1 mRNA levels") + ylab("Gene ANXA1 Protein Copy levels")
# 
# cor(x = protein_and_rna[, 2], y = protein_and_rna[, 22], method = "spearman")


```


